A Lightweight AI-based Approach For Drone Jamming Detection

### 2024-10-13 article.tex
Revisions pt.6

% Abstract (Do not insert blank lines, i.e. \\) 
\abstract{The future integration of drones in 6G networks will significantly enhance their capabilities, enabling a wide range of new applications based on autonomous operation. However, drone networks are particularly vulnerable to Jamming attacks, a type of Availability attack that can disrupt network operation and hinder drone functionality. In this paper, we propose an unsupervised machine learning approach for the detection of constant and periodic jamming attacks, using the Isolation Forest algorithm. We detail the tuning of the base model as well as the intregration with a Majority Rule module which significantly reduced the number of false positives caused by environmental noise, achieving high accuracy and precision. Our approach outperforms the standard Isolation Forest model in the detection of both constant and periodic jamming attacks, while still correctly identifying normal traffic. Finally, we discuss the potential integration of the proposed solution in 6G-enabled drone networks, as a lightweight edge-based solution for enhancing security against jamming attacks. }

% Keywords
\keyword{Jamming attacks, 6G Drone Networks, Isolation Forest, Machine Learning, Edge AI, IoT Security}

@@

The shift from 4G to 5G has been a generational leap has revolutionized connectivity. 
Despite 5G being still in its early adoption rate at the time of writing \cite{5GStatisticsTaylor}, the research community is already working on the next generation of wireless communication technologies, 6G.

6G technology is expected to enable a wide variety of new use cases, thanks to the increases in both data rates and latency, but this in turn 
will also bring forth new security challenges specific to those applications.

One of the main differences between 5G and 6G technology will be a much deeper integration with Artificial Intelligence (AI). While Software Defined Networks (SDN) have played a key role in improving the efficiency and security 
of 5G networks, 6G is expected to take this a step further, by designing the networks to be integrated with AI from the ground up. This is what the authors of \cite{6GRoadmapLetaief} define as 
the shift from \textit{Softwarization} to \textit{Intelligentization}.

AI integration in 6G networks will greatly strengthen the security of the network against potential threats. 
By leveraging Diagnostic Analytics, a collection of insights into the status of the networks, security teams 
will be able to train specific AI models to detect and respond to security threats in real-time.

In this paper we will focus on a specific aspect of the security of 6G networks, namely we will provide a lightweight edge-based machine learning approach 
for the detection of jamming attacks in networks of drones.

We conceptualize a scenario where a drone is subject to constant and periodic jamming and has to rely on an internal AI model to detect the attack and apply the appropriate countermeasures.

In the future, IoT devices will be equipped with AI-specific chips that will allow them to run AI models directly on the edge, without the need to offload computation to a central server. This is especially useful in the case of jamming attacks, where the communication between the device and the server is disrupted.

Our solution is designed to run on a resource constrained IoT device like a drone and to be lightweight, both in terms of memory usage and computational power.
Despite the low resources requirements, the proposed approach is able to detect both constant and periodic jamming attacks with a high degree of accuracy, as well as correctly classify normal traffic.

The paper is divided in 5 sections: 
\begin{enumerate}
	\item \textbf{Topic overview:} General overview of the benefits of drone integration in 6G networks, analysis of jamming attacks: types, mitigation and detection. Discussion of the advantages provided by an edge AI approach for jamming detection. 
	\item \textbf{System Model:} Description of the analyzed scenario, along with the dataset, algorithms, and evaluation metrics chosen for the tests.
	\item \textbf{Model Performance:} Explanation of the testing methodology, including model tuning, testing phases, and detailed description of the performance of the model.
	\item \textbf{Discussion:} Analysis of the results from the previous section, emphasizing the advantages of the proposed approach.
	\item \textbf{Conclusions:} Overview of how the proposed solution applies to real-world scenarios, with a discussion on potential future research directions.
\end{enumerate}

@@

This work build upon the research done in \cite{JammingDetectionIoT-Hussain}, where the authors analyze a machine learning based approach for the detection of jamming attacks. 
The authors propose the use of Convolutional Neural Networks (CNNs) to detect jamming attacks in a IoT drone scenario. The authors \cite{JammingDetectionIoT-Hussain} show that their CNN model is able to achieve a high detection rate against constant jamming attacks, but
struggles in the classification of periodic jamming and normal traffic.

@@

In the military sector, the effectiveness of drones is widely recognized, both for high and low end models and are being used in a variety of roles, from more passive roles like surveillance and intelligence gathering, to more active roles like
delivery of explosives and targeted strikes.

Securing the communication channel against Jamming attacks is crucial in all these applications, especially in a safety critical environment.

@@

Jamming attacks are a type of Denial of Service (DoS) attack that aims at disrupting the physical communication between two or more devices. 
This is achieved by transmitting a signal on the same frequency as the one used by the devices to communicate. If the jamming signal power level is high enough, 
it is able to overwhelm the legitimate signal, effectively blocking the communication between the devices \cite{DroneCommHassija}. 
Since the ability to communicate is affected, Jamming attacks falls under the umbrella of attacks that target the \textit{Availability} of the service in the CIA triad \cite{DataIntegrityCawthra}. 
Jamming attacks can be classified into 5 main categories, based on the attack pattern of the jammer:

\begin{itemize}
    \item \textbf{Constant Jamming:} The jammer continuously transmits a strong signal on the same frequency used by the devices it wants to disrupt to communicate. 
    \item \textbf{Periodic Jamming:} The jammer transmits a strong signal for a certain period of time \(t_a\), then stops transmitting for another period of time \(t_b\). This cycle is repeated until the attack ends.
    \item \textbf{Random Jamming:} In random jamming, the jammer is active at random intervals, jamming each transmitted packet with a probability 
    \(p\) based on a random pattern \cite{VANETsAI-Lyamin}. 
    \item \textbf{Reactive Jamming:} A reactive jammer starts transmitting its jamming signal only when it senses energy in the communication channel, 
    indicating that a legitimate transmission is taking place. This type of jamming attack is more power efficient compared to other 
    attack patterns, as it only transmits its signal when it know that it can actually disrupt the communication \cite{MLMisbehavior5GBoualouache}.
    \item \textbf{Smart Jamming:} A smart jammer is a more sophisticated type of jammer that is able to adapt its jamming signal to 
    maximize the disruption of the communication between the devices. Smart jammers employ a traffic analysis module, meaning they are able to modify their attack pattern based on the 
    transmission specifics of the devices they are targeting and adapt to changes in the communication channel \cite{AntiJammingV2V-Feng}.
\end{itemize}

@@

Jamming attacks are particularly effective against drone networks, as drones usually rely on external input to navigate and operate correctly.
If a jammer were able to completely block the communication between the drone and the base station, the drone would be left without any indication on how to behave and would need to activate an internal failsafe mechanism. 
This usually comes in the form of either a \textit{return to base} procedure, a \textit{hover in place} procedure or a \textit{land in place} procedure.
All of these approaches leave the drone in a vulnerable position, as a bad actor could potentially capture the drone and use it for malicious purposes.
This is especially true when jamming attacks are used in combination with other types of attacks, such as spoofing attacks. A jamming attack can be initiated against a drone, making it lose connection with the central server. 
The drone would be forced to start a failsafe procedure, like \textit{return to base}. The attacker could than employ a GPS spoofing attack to make the drone believe that the base station is in a different location, leading to the drone landing in the spot designated by the attacker \cite{RQ170DroneOwano}

@@

When presenting a Machine Learning (ML) based approach in an IoT setting, the question of where the AI model should be placed often arises. By their nature, IoT devices, and in turn drones, are usually resource constrained, 
both in terms of computational power but also in terms of internal storage and battery capacity \cite{6GSecurity-Chorti}. This means that complex AI models and algorithms are usually not feasible to be run on the device itself. A centralized approach 
offloads the computational burden to a central server, that returns the results of the AI model to the device. While this scheme might be favorable in some cases, in a real-time situation such as a jamming attack, the latency introduced by the
communication with the server means less time to react to the attack. Also, as jamming attacks degrade or sometimes completely block the communication between the device and the base station, a centralized approach might not be feasible in this case.
A collaborative edge approach would suffer from the same problems, as there is a high likelihood that the communication between the devices would be disrupted by the attack. 
Implementing a lightweight on-device edge approach, on the other hand, while not as precise as a centralized approach, would provide real-time results and would be able to operate even when the communication is disrupted, as is the case in a jamming situation.

@@

In the proposed scenario we have a flying drone that is subject to a jamming attack. The drone communicates with a ground station and periodically samples the Received Signal Strength (RSS) of the signal it receives. 
The drone implements a simple unsupervised machine learning algorithm module, trained on nominal traffic RSS values. The module takes the sampled RSS values and classifies them as either normal or anomalous, making the drone able to determine wether or not a jamming attack is taking place.

We assume that a jammer is deployed in the network and that it starts a jamming attack at a certain instant by transmitting a high power signal on the same frequency as the one used by the drone to communicate with the ground station.
We assume that the jammer is able to completely block the communication between the drone and the ground station, meaning the drone has to rely on its internal classification algorithm to determine if a jamming attack is taking place and react accordingly.
We assume that the jammer type is unknown to the drone.

Figure \ref{fig:CombinedJammingscenariosDiagram} shows a diagram of the proposed scenario.

\begin{figure}[H]
	\includegraphics[width=10.5 cm]{img/CombinedJammingscenariosDiagram.jpg}
	\caption{Proposed jamming scenario. The Drone communicates with the 6G base station and receives information about how to behave. The fixed jammer \textbf{(1)} and the mobile jammer \textbf{(2)} target the communication between the drone and the base station. The drone has an internal classification module that is able to detect jamming attacks.}
	\label{fig:CombinedJammingscenariosDiagram}
	\end{figure}

In our scenario, the drone will be subject to two types of jamming attacks, a constant and a periodic jamming attack.
These types of attacks can be attributed to different types of jammers. The constant jamming attack could be caused by a ground jammer, as maintaining a constant jamming signal requires a large power source, while 
the periodic jamming attack could be caused by a mobile jammer, for example a drone, as drones are usually battery powered and can only jam for a limited amount of time. 
Employing a periodic jamming attack could help the malicious drone preserve energy and reduce the chances of being detected.

Table \ref{tab:attacker_classification} shows the attacker classification in the proposed scenario \cite{MLMisbehavior5GBoualouache}:

@@

The dataset contains 3 different \textit{.txt} files that store the RSS values sampled by the Raspberry Pi 3. The first file contains the RSS values sampled during a constant jamming attack, the second file contains the RSS values sampled during a periodic jamming attack and the third file contains the RSS values sampled during normal network operation.
The samples are stored in a single column, with each row representing a single sampled RSS value, represented using the \textit{dBm (decibel-milliwatts)} unit of measurement.

@@

When choosing the algorithm for the classification module, we first had to define the requirements that the algorithm has to meet. The algorithm has to be able to classify anomalous samples with a high detection rate (Recall), while at the same time 
being lightweight enough in terms of memory usage and computational power to be run on a resource constrained device like a drone. The training phase should also be fast, as this would allow the drone to quickly determine the normal transmission RSS values in a constantly changing environment. A periodical redefinition of normal RSS values would 
limit the number of false positives caused by the drone mobility. 
The model should be also required a small amount of storage to run effectively, as drones are usually very limited in internal memory.

The algorithm that best fit these requirements was the \textit{Isolation Forest} algorithm \cite{IsolationForestLiu}. The Isolation Forest (IF) algorithm is a state-of-the-art unsupervised machine learning algorithm for anomaly detection, introduced by Liu et al. in 2008. 
Unlike traditional anomaly detection algorithms, which require the definition of a normal class, IF is able to detect anomalies without explicitly defining their characteristics.
This is achieved by leveraging the properties of anomalies themselves, i.e being rare and separated from the majority of the data points.

@@

\begin{figure}[H]
	\includegraphics[width=6 cm]{img/IsolationForest.jpg}
	\caption{Isolation forest in action. The anomaly \textbf{x0} is isolated in less splits compared to the normal datapoint \textbf{x1} \cite{IsolationForestLiu}.}
	\label{fig:IsolationForest}
\end{figure}   
\unskip

@@

\begin{itemize}
	\item \textbf{Accuracy:} the ratio of correctly classified data points to the total number of data points.
	\begin{equation}
		\label{eq:accuracy}
	  Accuracy = \frac{TP + TN}{TP + TN + FP + FN}
	\end{equation}
	\item \textbf{Precision:} the ratio of correctly classified anomalies to the total number of data points classified as anomalies.
	\begin{equation}
		\label{eq:precision}
	  Precision = \frac{TP}{TP + FP}
	\end{equation}
	\item \textbf{Recall:} the ratio of correctly classified anomalies to the total number of anomalies.
	\begin{equation}
		\label{eq:recall}
	  Recall = \frac{TP}{TP + FN}
	\end{equation}
	\item \textbf{F1 Score:} the harmonic mean of the precision and recall.
	\begin{equation}
		\label{eq:f1}
	  F1 = 2 \times \frac{Precision \times Recall}{Precision + Recall}
	\end{equation}
\end{itemize}

The combination of these metrics gives us concrete insight into the performance of the model.

@@

The first phase in our testing methodology was the tuning of the model's hyperparameters. The tuning was performed by making the parameters vary between a minimum and a maximum value, evaluating the impact on the performance metrics (detailed in Section \ref{EvaluationMetrics}) and then choosing the best values.
Among the parameters available in the \textit{Scikit-learn} implementation of the Isolation Forest, we decided to tune the following:

\begin{itemize}
	\item \textbf{n\_estimators:} the number of base estimators in the ensamble, i.e the number of isolation trees used to compute the anomaly score of each datapoint. 
	\item \textbf{max\_samples:} the max number of samples to draw from the dataset to train each tree.
	\item \textbf{contamination:} the amount of outliers present in the training dataset.  
\end{itemize}

During the tuning phase, the untested hyperparameters were set to the default values of the \textit{Scikit-learn} implementation of the Isolation Forest (table \ref{tab:isolation_forest_parameters}).

@@

The values whe chose for \textit{normal\_traffic\_size} and \textit{jamming\_size} are shown in Table \ref{tab:dataset_sizes}.

@@

The dataset is intentionally unbalanced between normal data points and anomalous points, as jamming attacks are usually rare events compared to normal network operation.

In Figure \ref{fig:InputSignal} we can see a representation of the input signal in the case of a constant jamming attack. The proposed results, unless otherwise specified, are based on the input signal shown in Figure \ref{fig:InputSignal}, which employs constant jamming. Periodic jamming always showed comparable trends as constant jamming in all the performed tests.

\begin{figure}[H]
    \begin{adjustwidth}{-\extralength}{0cm}
        \centering
        \includegraphics[width=21cm]{img/InputSignal.png}
    \end{adjustwidth}
    \caption{Input signal in the case of Constant Jamming. The normal traffic signal is concatenated with the jamming signal.}
    \label{fig:InputSignal}
\end{figure}

@@

\begin{figure}[H]
	\includegraphics[width=10.5cm]{img/nEstimatorsTuning.png}
	\caption{n\_estimators tuning results.}\label{fig:n_estimators_tuning}
\end{figure}
\unskip

@@

From Figure \ref{fig:max_samples_tuning} we can see that the model performance is not greatly affected by the \textit{max\_samples} parameter. This is most likely due to the great difference in RSS values between the normal traffic samples and the jamming attack samples, meaning the model can develop the ability to correctly classify the input even with a limited training set.

\begin{figure}[H]
	\includegraphics[width=10.5cm]{img/maxSamplesTuning.png}
	\caption{max\_samples tuning results.}\label{fig:max_samples_tuning}
\end{figure}
\unskip

@@

The final parameter we focused on is the \textit{contamination} parameter. The tuning values are shown in Table \ref{tab:contamination_tuning}.

@@

\begin{figure}[H]
	\includegraphics[width=10.5cm]{img/contaminationTuning.png}
	\caption{contamination tuning results.}\label{fig:contamination_tuning}
\end{figure}
\unskip

Our testing showed that, as expected, tuning of the contamination parameter has minimal impact on the training and classification time.

@@

The tuning phase proved effective in improving the model's performance compared to the default hyperparameters,with an increase of $2\%$ in accuracy, $6\%$ in precision, and $5.1\%$ in the F1 score, while maintaining a nearly identical recall ($0.1\%$ drop). However, despite reaching a very high Recall score in both scenarios, the model is still too prone to false positives, as indicated by the value of the Precision metric. 
This is most likely due to the nature of the normal traffic signal. As we see from Figure \ref{fig:standardIsolationResults}, many of the normal traffic samples have RSS values that are similar to the jamming attack samples, leading the model to classify them as anomalies.
This is most likely because of how the dataset has been created, as the normal traffic samples were measured in a real-world scenario where the RSS values would be effected by external noise.

@@

\subsubsection{Majority rule model testing} \label{sec:majorityRuleModelTesting}

Having understood the nature of the problem, we decided to implement a majority rule system to reduce the number of false positives. The improved model would use the Tuned Isolation Forest model's results and then pass them to a Majority Rule module, as shown in Figure \ref{fig:MajorityRuleDiagram}.

\begin{figure}[H]
	\includegraphics[width=12.5 cm]{img/ModulesStructure.jpg}
	\caption{Majority Rule system diagram. The Tuned Isolation Forest model results are passed to the Majority Rule module, that then provides the final classification.}
	\label{fig:MajorityRuleDiagram}
	\end{figure}

The logic behind the Majority Rule module is simple. The Majority Rule takes as input the classification results of the IF model and inserts them into a sliding window. If the datapoint is classified by the Tuned IF as an anomaly, the Majority Rule module checks the contents of the sliding window. If the 
majority of the datapoints in the window are classified as anomalies, the data point is classified as an anomaly. If the majority of the datapoints in the window are classified as normal, the data point is classified as normal (see code snipped \ref{alg:majority_rule}).

@@

\begin{table}[H]
	\caption{window\_size tuning values.}\label{tab:window_size_tuning}
	\begin{tabularx}{\textwidth}{CC}
	\toprule
	\textbf{window\_size} & \textbf{Values} \\
	\midrule
	Minimum & 1 \\
	Maximum & 100 \\
	Step & 1 \\
	\bottomrule
\end{tabularx}
\end{table}

@@

\bibitem[Boualouache and Engel(2023)]{MLMisbehavior5GBoualouache}
Boualouache, A.; Engel, T. A Survey on Machine Learning-Based Misbehavior Detection Systems for 5G and Beyond Vehicular Networks. {\em IEEE Communications Surveys and Tutorials} {\bf 2023}, {\em 25}, 1128--1172. https://doi.org/10.1109/COMST.2023.3236448.

\bibitem[Sobot(2012)]{WirelessCommSobot}
Sobot, R. \textit{Wireless Communication Electronics: Introduction to RF Circuits and Design}; Springer: 2012; p. 252. ISBN 9783030486303.

>>>>

% Abstract (Do not insert blank lines, i.e. \\) 
\abstract{The future integration of drones in 6G networks will significantly enhance their capabilities, enabling a wide range of new applications based on autonomous operation. However, drone networks are particularly vulnerable to Jamming attacks, a type of Availability attack that can disrupt network operation and hinder drone functionality. In this paper, we propose a low complexity unsupervised machine learning approach for the detection of constant and periodic jamming attacks, using the Isolation Forest algorithm. We detail the tuning of the base model as well as the intregration with a Majority Rule module which significantly reduced the number of false positives caused by environmental noise, achieving high accuracy and precision. Our approach outperforms the standard Isolation Forest model in the detection of both constant and periodic jamming attacks, while still correctly identifying nominal traffic. Finally, we discuss the potential integration of the proposed solution in 6G-enabled drone networks, as a lightweight edge-based solution for enhancing security against jamming attacks. }

% Keywords
\keyword{Jamming attacks, 6G Drone Networks, Isolation Forest, Edge AI, IoT Security}

@@

The shift from 4G to 5G has been a generational leap has revolutionized connectivity. 
Despite 5G being still in its early adoption rate at the time of writing \cite{5GStatisticsTaylor}, the research community is already working on the next generation of wireless communication technologies, 6G. 
6G technology is expected to enable a wide variety of new use cases, thanks to massive coverage as well as increases in both data rates and latency. All of these advancements will in turn bring forth new security challenges specific to those applications.

One of the main differences between 5G and 6G technology will be a much deeper integration with Artificial Intelligence (AI). While Software Defined Networks (SDN) have played a key role in improving the efficiency and security 
of 5G networks, 6G is expected to take this a step further, by designing the networks to be integrated with AI from the ground up. This is what the authors of \cite{6GRoadmapLetaief} define as 
the shift from \textit{Softwarization} to \textit{Intelligentization}.AI integration in 6G networks will greatly strengthen the security of the network against potential threats. 
By leveraging Diagnostic Analytics, a collection of insights into the status of the networks, security teams 
will be able to train specific AI models to detect and respond to security threats in real-time.

In this paper we focus on a specific aspect of the security of 6G networks, namely we provide a lightweight edge-based Machine Learning approach 
for the detection of jamming attacks in networks of drones. We conceptualize a scenario where a drone is subject to constant and periodic jamming and has to rely on an internal AI model to detect the attack and apply the appropriate countermeasures.

In the future, IoT devices are expected to come equipped with AI-specific chips that will allow them to run AI models directly on the edge, without the need to offload computation to a central server. This is especially useful in the case of jamming attacks, where the communication between the device and the server is disrupted.
Our solution is designed to run on a resource constrained IoT device like a drone and to be lightweight, both in terms of memory usage and computational power.
Despite the low resources requirements, the proposed approach is able to detect both constant and periodic jamming attacks with a high degree of accuracy, as well as correctly classify normal traffic.

The paper is organized in 5 Sections: 
\begin{enumerate}
	\item \textbf{Topic overview:} General overview of the benefits of drone integration in 6G networks, analysis of jamming attacks: types, mitigation and detection. Discussion of the advantages provided by an edge AI approach for jamming detection. 
	\item \textbf{System Model:} Description of the analyzed scenario, along with the dataset, algorithms, and evaluation metrics chosen for the tests.
	\item \textbf{Model Performance:} Explanation of the testing methodology, including model tuning, testing phases, and detailed description of the performance of the model.
	\item \textbf{Discussion:} Analysis of the results from the previous section, emphasizing the advantages of the proposed approach.
	\item \textbf{Conclusions:} Overview of how the proposed solution applies to real-world scenarios, with a discussion on potential future research directions.
\end{enumerate}

@@

This work build upon the research done in \cite{JammingDetectionIoT-Hussain}, where the authors analyze a Machine Learning (ML) based approach for the detection of jamming attacks. 
The authors propose the use of Convolutional Neural Networks (CNNs) to detect jamming attacks in a IoT drone scenario. The authors \cite{JammingDetectionIoT-Hussain} show that their CNN model is able to achieve a high detection rate against constant jamming attacks, but
struggles in the classification of periodic jamming and normal traffic.

@@

In the military sector, the effectiveness of drones is widely recognized, both for high and low end models and are being used in a variety of roles, from more passive roles like surveillance and intelligence gathering, to more active roles like
delivery of explosives and targeted strikes. Securing the communication channel against jamming attacks is crucial in all these applications, especially in a safety critical environment.

@@

Jamming attacks are a type of Denial of Service (DoS) attack that aims at disrupting the physical communication between two or more devices. 
This is achieved by transmitting a signal on the same frequency as the one used by the devices to communicate. If the jamming signal power level is high enough, 
it is able to overwhelm the legitimate signal, effectively blocking the communication between the devices \cite{DroneCommHassija}. 
Since the ability to communicate is affected, jamming attacks falls under the umbrella of attacks that target the \textit{Availability} of the service in the CIA triad \cite{DataIntegrityCawthra}. 
Jamming attacks can be classified into 5 main categories, based on the attack pattern of the jammer:

\begin{itemize}
    \item \textbf{Constant Jamming:} The jammer continuously transmits a strong signal on the same frequency used by the devices it wants to disrupt to communicate. 
    \item \textbf{Periodic Jamming:} The jammer transmits a strong signal for a certain period of time \(t_{on}\), then stops transmitting for another period of time \(t_{off}\). This cycle is repeated until the attack ends.
    \item \textbf{Random Jamming:} In random jamming, the jammer is active at random intervals, jamming each transmitted packet with a probability 
    \(p\) based on a random pattern \cite{VANETsAI-Lyamin}. 
    \item \textbf{Reactive Jamming:} A reactive jammer starts transmitting its jamming signal only when it senses energy in the communication channel, 
    indicating that a legitimate transmission is taking place. This type of jamming attack is more power efficient compared to other 
    attack patterns, as it only transmits its signal when it know that it can actually disrupt the communication \cite{MLMisbehavior5GBoualouache}.
    \item \textbf{Smart Jamming:} A smart jammer is a more advanced and computationally intensive type of jammer that is able to adapt its jamming signal to 
    maximize the disruption of the communication between the devices. Smart jammers employ a traffic analysis module, meaning they are able to modify their attack pattern based on the 
    transmission specifics of the devices they are targeting and adapt to changes in the communication channel \cite{AntiJammingV2V-Feng}. 
\end{itemize}

@@

Jamming attacks are particularly effective against drone networks, as drones usually rely on external input to navigate and operate correctly.
If a jammer were able to completely block the communication between the drone and the Base Station (BS), the drone would be left without any indication on how to behave and would need to activate an internal failsafe mechanism. 
This usually comes in the form of either a \textit{return to base} procedure, a \textit{hover in place} procedure or a \textit{land in place} procedure.
All of these approaches leave the drone in a vulnerable position, as a bad actor could potentially capture the drone and use it for malicious purposes.
This is especially true when jamming attacks are used in combination with other types of attacks, such as spoofing attacks. A jamming attack can be initiated against a drone, making it lose connection with the central server. 
The drone would be forced to start a failsafe procedure, like \textit{return to base}. The attacker could than employ a GPS spoofing attack to make the drone believe that the Base Station is in a different location, leading to the drone landing in the spot designated by the attacker \cite{RQ170DroneOwano}

@@

When presenting a ML based approach in an IoT setting, the question of where the AI model should be placed often arises. By their nature, IoT devices, and in turn drones, are usually resource constrained, 
both in terms of computational power but also in terms of internal storage and battery capacity \cite{6GSecurity-Chorti}. This means that complex AI models and algorithms are usually not feasible to be run on the device itself. A centralized approach 
offloads the computational burden to a remote central server, that returns the results of the AI model to the device. While this scheme might be favorable in some cases, in a real-time situation such as a jamming attack, the latency introduced by the
communication with the server reduces the time to react to the attack. Also, as jamming attacks degrade or sometimes completely block the communication between the device and the BS, a centralized approach might not be feasible in this case.
A collaborative edge approach would suffer from the same problems, as there is a high likelihood that the communication between the devices would be disrupted by the attack. 
Implementing a lightweight on-device edge approach, on the other hand, while not as precise as a centralized approach, would provide real-time results and would be able to operate even when the communication is disrupted, as is the case in a jamming situation.

@@

In the proposed scenario (Figure \ref{fig:CombinedJammingscenariosDiagram}) we have a flying drone that is subject to a jamming attack. The drone communicates with a ground station and periodically samples the Received Signal Strength (RSS) of the signal it receives. 
The drone implements a simple unsupervised machine learning algorithm module, trained on nominal traffic RSS values. The module takes the sampled RSS values and classifies them as either normal or anomalous, making the drone able to determine wether or not a jamming attack is taking place.

We assume that a jammer is deployed in the network and that it starts a jamming attack at a certain instant by transmitting a high power signal on the same frequency as the one used by the drone to communicate with the ground station.
We assume that the jammer is able to completely block the communication between the drone and the ground station, meaning that the drone has to rely on its internal classification algorithm to determine if a jamming attack is taking place and react accordingly.
We assume that the jammer type is unknown to the drone.

\begin{figure}[H]
	\includegraphics[width=10.5 cm]{img/CombinedJammingscenariosDiagram.jpg}
	\centering
	\caption{Proposed jamming scenario. The Drone communicates with the 6G BS and receives information about how to behave. The fixed jammer \textbf{(1)} and the mobile jammer \textbf{(2)} target the communication between the drone and the Base Station. The drone has an internal classification module that is able to detect jamming attacks.}
	\label{fig:CombinedJammingscenariosDiagram}
	\end{figure}

In our scenario, the drone is subject to two types of jamming attacks, a constant and a periodic jamming attack that can be attributed to two different types of jammers. The constant jamming attack could be caused by a fixed ground jammer, as maintaining a constant jamming signal requires a large power source, while 
the periodic jamming attack could be caused by a mobile jammer, for example a drone, as drones are usually battery powered and can only jam for a limited amount of time. 
Employing a periodic jamming attack could help the malicious drone preserve energy and reduce the chances of being detected.

The attacker classification \cite{MLMisbehavior5GBoualouache} of the proposed scenario is detailed in Table \ref{tab:attacker_classification}:

@@

The dataset contains 3 different \textit{.txt} files that store the RSS values sampled by the Raspberry Pi 3. The first file contains the RSS values sampled during a constant jamming attack, the second file contains the RSS values sampled during a periodic jamming attack and the third file contains the RSS values sampled during normal network operation.
The samples are stored in a single column, with each row representing a single sampled RSS value, expressed in \textit{dBm (decibel-milliwatts)}.

@@

When choosing the algorithm for the classification module, we first had to define the requirements that the it is required to meet. The algorithm has to have the ability to classify anomalous samples with a high detection rate (Recall), while at the same time 
being lightweight enough in terms of memory usage and computational power to be run on a resource constrained device like a drone. The training phase should also be fast, as this would allow the drone to quickly determine the normal transmission RSS values in a constantly changing environment. A periodical redefinition of normal RSS values would 
limit the number of false positives caused by the drone mobility. 
The model should be also required a small amount of storage to run effectively, as drones are usually very limited in internal memory.

The algorithm that we deemed the best fit for these requirements is the \textit{Isolation Forest} algorithm \cite{IsolationForestLiu}. The Isolation Forest (IF) algorithm is a state-of-the-art unsupervised ML algorithm for anomaly detection, introduced by Liu et al. in 2008. 
Unlike traditional anomaly detection algorithms, which require the definition of a normal class, IF is able to detect anomalies without explicitly defining their characteristics.
This is achieved by leveraging the properties of anomalies themselves, i.e being rare and separated from the majority of the data points.

@@

\begin{figure}[H]
	\centering
	\includegraphics[width=6 cm]{img/IsolationForest.jpg}
	\caption{Isolation forest in action. The anomaly \textbf{x0} is isolated in less splits compared to the normal datapoint \textbf{x1} \cite{IsolationForestLiu}.}
	\label{fig:IsolationForest}
\end{figure}   
\unskip

@@

\begin{itemize}
	\item \textbf{Accuracy:} the ratio of correctly classified data points to the total number of data points.
	\begin{equation}
		\label{eq:accuracy}
	  Accuracy = \frac{TP + TN}{TP + TN + FP + FN}
	\end{equation}
	\item \textbf{Precision:} the ratio of correctly classified anomalies to the total number of data points classified as anomalies.
	\begin{equation}
		\label{eq:precision}
	  Precision = \frac{TP}{TP + FP}
	\end{equation}
	\item \textbf{Recall:} the ratio of correctly classified anomalies to the total number of anomalies.
	\begin{equation}
		\label{eq:recall}
	  Recall = \frac{TP}{TP + FN}
	\end{equation}
	\item \textbf{F1 Score:} the harmonic mean of the precision and recall.
	\begin{equation}
		\label{eq:f1}
	  F1 = 2 \times \frac{Precision \times Recall}{Precision + Recall}
	\end{equation}
\end{itemize}
The combination of these metrics gives us concrete insight into the performance of the model.

@@

The first phase in our testing methodology is the tuning of the model's hyperparameters by making the parameters vary between a minimum and a maximum value, evaluating the impact on the performance metrics (detailed in Section \ref{EvaluationMetrics}) and then choosing the best values.
Among the parameters available in the \textit{Scikit-learn} implementation of the Isolation Forest, we decided to tune the following:

\begin{itemize}
	\item \textbf{n\_estimators:} the number of base estimators in the ensamble, i.e the number of isolation trees used to compute the anomaly score of each datapoint. 
	\item \textbf{max\_samples:} the max number of samples to draw from the dataset to train each tree.
	\item \textbf{contamination:} the amount of outliers present in the training dataset.  
\end{itemize} 
During the tuning phase, the untested hyperparameters were set to the default values of the \textit{Scikit-learn} implementation of the Isolation Forest (table \ref{tab:isolation_forest_parameters}).

@@

The selected values for \textit{normal\_traffic\_size} and \textit{jamming\_size} are shown in Table \ref{tab:dataset_sizes}.

@@

The dataset is intentionally unbalanced between normal data points and anomalous points, as jamming attacks are usually rare events compared to normal network operation. 
In Figure \ref{fig:InputSignal} we can see a representation of the input signal in the case of a constant jamming attack. The proposed results, unless otherwise specified, are based on the input signal shown in Figure \ref{fig:InputSignal}, which employs constant jamming. Periodic jamming always shown comparable trends as constant jamming in all the performed tests.

\begin{figure}[H]
    \begin{adjustwidth}{-\extralength}{0cm}
        \centering
        \includegraphics[width=21cm]{img/InputSignal.png}
    \end{adjustwidth}
    \caption{Input signal in the case of Constant jamming. The normal traffic signal is concatenated with the jamming signal.}
    \label{fig:InputSignal}
\end{figure}

@@

\begin{figure}[H]
	\centering
	\includegraphics[width=10.5cm]{img/nEstimatorsTuning.png}
	\caption{n\_estimators tuning results.}\label{fig:n_estimators_tuning}
\end{figure}
\unskip

@@

From Figure \ref{fig:max_samples_tuning} we can notice that the model performance is not greatly affected by the \textit{max\_samples} parameter. This is most likely due to the great difference in RSS values between the normal traffic samples and the jamming attack samples, meaning the model can develop the ability to correctly classify the input even with a limited training set.

\begin{figure}[H]
	\centering
	\includegraphics[width=10.5cm]{img/maxSamplesTuning.png}
	\caption{max\_samples tuning results.}\label{fig:max_samples_tuning}
\end{figure}
\unskip

@@

The last parameter we focused on is the \textit{contamination} parameter. The tuning values are shown in Table \ref{tab:contamination_tuning}.

@@

\begin{figure}[H]
	\centering
	\includegraphics[width=10.5cm]{img/contaminationTuning.png}
	\caption{contamination tuning results.}\label{fig:contamination_tuning}
\end{figure}
\unskip

Our testing pointed out that, as expected, tuning of the contamination parameter has minimal impact on the training and classification time.

@@

The tuning phase proved effective in improving the model's performance compared to the default hyperparameters, with an increase of $2\%$ in accuracy, $6\%$ in precision, and $5.1\%$ in the F1 score, while maintaining a nearly identical recall ($0.1\%$ drop). However, despite reaching a very high Recall score in both scenarios, the model is still too prone to false positives, as indicated by the value of the Precision metric. 
This is most likely due to the nature of the normal traffic signal. As we see from Figure \ref{fig:standardIsolationResults}, many of the normal traffic samples have RSS values that are similar to the jamming attack samples, leading the model to classify them as anomalies.
This is most likely because of how the dataset has been created, as the normal traffic samples were measured in a real-world scenario where the RSS values would be effected by external noise.

@@

\subsubsection{Majority Rule model testing} \label{sec:majorityRuleModelTesting}

Having understood the nature of the problem, we decided to implement a Majority Rule (MR) system to reduce the number of false positives. The improved model would use the Tuned Isolation Forest model's results and then pass them to a MR module, as shown in Figure \ref{fig:MajorityRuleDiagram}.

\begin{figure}[H]
	\centering
	\includegraphics[width=12.5 cm]{img/ModulesStructure.jpg}
	\caption{Majority Rule system diagram. The Tuned Isolation Forest model results are passed to the MR module, that then provides the final classification.}
	\label{fig:MajorityRuleDiagram}
	\end{figure}

The logic behind the Majority Rule module is simple. The MR takes as input the classification results of the IF model and inserts them into a sliding window. If the datapoint is classified by the Tuned IF as an anomaly, the MR module checks the contents of the sliding window. If the 
majority of the datapoints in the window are classified as anomalies, the data point is classified as an anomaly. If the majority of the datapoints in the window are classified as normal, the data point is classified as normal (see code snipped \ref{alg:majority_rule}).

@@

\begin{table}[H]
	\centering
	\caption{window\_size tuning values.}\label{tab:window_size_tuning}
	\begin{tabularx}{\textwidth}{CC}
	\toprule
	\textbf{window\_size} & \textbf{Values} \\
	\midrule
	Minimum & 1 \\
	Maximum & 100 \\
	Step & 1 \\
	\bottomrule
\end{tabularx}
\end{table}

@@
